Using language as a indicator of cognitive decline in pre-clinical Alheimer's Disease

### 2019-11-13 Source/Paper 1 - Presidents Corpus Paper/PresidentsCorpus.tex
updated abstract
updated discussion

\title{Exploring the Presidents Corpus}

@@

\keywords{Mild Cognitive Impairment \and Alzheimer's Disease \and Machine Learning \and Diagnosis \and Natural Language Processing} 
\bigskip
\begin{abstract}
Several research efforts aimed at improving the early diagnosis of Alzheimer's Disease have focused on the longitudinal analysis of lexical features which have been implicated in the pre-clinical progression of this disease. We build on these studies and continue to refine a method for extracting and analysing language from transcripts. This method is based on a statistical evaluation of the most relevant features, as well as on the use of Generalised Additive Models instead of Linear Regression as a way to more adequately model the declines in language over time. We evaluate this method using the unscripted question and answer portions of news conferences from three U.S Presidents: Ronald Reagan (RR), who would later go on to receive a diagnosis of Alzheimer's Disease; George H. W. Bush (GB) and Donald Trump (DT), who have no known diagnosis of dementia. Our analysis uncovered several lexical features that did not change significantly for GB and DT, but presented significant changes for RR, including significant decreases in unique words and noun usage, and significant increases in pronouns (particularly personal pronouns), conjunctions and words which are related to social processes significantly increase over time. These findings are consistent with the psychological literature, which documents a link between these features and cognitive decline in Alzheimer's Disease. This work points the way to developing a comparatively non-invasive protocol that could be used in clinical settings that help to spot atypical declines in language use over time, potentially contributing to earlier, more accurate diagnoses of Alzheimer's Disease.

\end{abstract}

@@

\section{Discussion}\label{discussion}
President Reagan received his diagnosis of AD in August 1994 but using transcripts of speeches he made in his two terms as President (January 1981 - January 1989) we have be able to identify certain changes in his use of language that we might ascribe to the onset of MCI and early AD. Despite differences in our methodology, our research supports some of the findings of Berisha and Liss in that we find an increase in non-specifc noun usage. Compared to our controls (GWHB and DJT), we find some slight trends with GWHB but no such trends with DJT in his speech albeit his samples of speech span a shorter amount of time.
\par 
A criticism of Berisha and Liss's work is the problems they had with normalising the transcripts in terms of length. This was also a problem in the work of Garrard et al \cite{Garrard2005, Le2011}. Whilst it is important to control for outliers, there are other ways in which we can control for length of sample. In this paper, we controlled for transcript length by dividing any features that were raw counts by the total length of the transcript. Given the increased amounts of hypothesis testing, it became necessary to control for false discoveries. 
\par 
Interestingly, when we normalised the various types of words used by the presidents we found some interesting patterns that further differentiated RR from the controls. Whilst Non-specific nouns increased over time, we found that noun usage in general significantly decreased and pronouns increased similarly significantly. The increase in pronoun for those with early AD has been identified in literature, although there are only a few studies that explore this \cite{Almor1999, Wendelstein2015}. Wendlestein et al propose that the increased used of pronouns is an expression of an impaired ability to adapt language to the listener's needs \cite{Wendelstein2015}. Almor et al attributed this reliance on pronouns due to a impaired working memory \cite{Almor1999}.
\par 
The decrease in overall noun usage has also been identified as a feature. Jarrold et al found that AD patients would use more pronouns, verbs and fewer nouns than controls \cite{Jarrold2014}. Wendlestein in their investigations into noun usage found that decreased later on in AD progression and was unaffected in the pre-clinical stages of AD \cite{Wendelstein2014}. Our results are supported by existing literature and this potentially means that language analysis in the way we have structured it may have diagnostic or prognostic properties.
One of the most striking differences between this study and the original is when looking at Unique Words. Whilst Unique Words as a feature is significant, which was a finding in the original study, when we controlled via normalisation, we found that this was not significant. 
\par
Another observation is that when conducting fitting a linear model to this data does not tell the whole story. For example, consider the following two plots. One shows a model fit to a linear  model, and the other shows a model fit using local polynomial regression model which aims to model the data more closely.

@@

\section{Conclusions}\label{conclusions}
The results of this work show that we can track a person's use of language through time in a number of ways, and it is possible for an individual to be his or her own control. This is important as it means the heterogenous nature of the MCI population does not impact results as much as if we were comparing those with MCI to controls. Equally, it would be helpful to have controls to ascertain what would be usual to expect in the decline of language in a healthy older adult. 
\par 
The results of this work also identify some clues as to what

>>>>

\title{Using language as a indicator of cognitive decline in pre-clinical Alheimer's Disease}

@@

\keywords{Mild Cognitive Impairment \and Alzheimer's Disease \and Machine Learning \and Diagnosis \and Natural Language Processing} 
\bigskip
\begin{abstract}
Several research efforts aimed at improving the early diagnosis of Alzheimer's Disease have focused on the longitudinal analysis of lexical features which have been implicated in the pre-clinical progression of this disease. We build on these studies and continue to refine a method for extracting and analysing language from transcripts. This method is based on a statistical evaluation of the most relevant features, as well as on the use of Generalised Additive Models instead of Linear Regression as a way to more adequately model the declines in language over time. We evaluate this method using the unscripted question and answer portions of news conferences from three U.S Presidents: Ronald Reagan (RR), who would later go on to receive a diagnosis of Alzheimer's Disease; George H. W. Bush (GB) and Donald Trump (DT), who have no known diagnosis of dementia. Our analysis uncovered several lexical features that did not change significantly for GB and DT, but presented significant changes for RR, including significant decreases in unique words and noun usage, and significant increases in pronouns (particularly personal pronouns), conjunctions and words which are related to social processes significantly increase over time. These findings are consistent with the psychological literature, which documents a link between these features and cognitive decline in Alzheimer's Disease. This work points the way to developing a comparatively non-invasive protocol that could be used in clinical settings that help to spot atypical declines in language use over time, potentially contributing to earlier, more accurate diagnoses of Alzheimer's Disease.
\end{abstract}

@@

\section{Discussion}\label{discussion}
President Reagan received his diagnosis of AD in August 1994 but using transcripts of speeches he made in his two terms as President (January 1981 - January 1989) we have be able to identify certain changes in his use of language that we might ascribe to the onset of MCI and early AD. Despite differences in our methodology, our research supports some of the findings of Berisha and Liss in that we find an increase in non-specifc noun usage. Compared to our controls (GWHB and DJT), we find some slight trends with GWHB but no such trends with DJT in his speech albeit his samples of speech span a shorter amount of time.
\par 
A criticism of Berisha and Liss's work is the problems they had with normalising the transcripts in terms of length. This was also a problem in the work of Garrard et al \cite{Garrard2005, Le2011}. Whilst it is important to control for outliers, there are other ways in which we can control for length of sample. In this paper, we controlled for transcript length by dividing any features that were raw counts by the total length of the transcript. Given the increased amounts of hypothesis testing, it became necessary to control for false discoveries. 
\par 
Interestingly, when we normalised the various types of words used by the presidents we found some interesting patterns that further differentiated RR from the controls. Whilst Non-specific nouns increased over time, we found that noun usage in general significantly decreased and pronouns increased similarly significantly. The increase in pronoun for those with early AD has been identified in literature, although there are only a few studies that explore this \cite{Almor1999, Wendelstein2015}. Wendlestein et al propose that the increased used of pronouns is an expression of an impaired ability to adapt language to the listener's needs \cite{Wendelstein2015}. Almor et al attributed this reliance on pronouns due to a impaired working memory \cite{Almor1999}.
\par 
The decrease in overall noun usage has also been identified as a feature. Jarrold et al found that AD patients would use more pronouns, verbs and fewer nouns than controls \cite{Jarrold2014}. Wendlestein in their investigations into noun usage found that decreased later on in AD progression and was unaffected in the pre-clinical stages of AD \cite{Wendelstein2014}. Our results are supported by existing literature and this potentially means that language analysis in the way we have structured it may have diagnostic or prognostic properties.
One of the most striking differences between this study and the original is when looking at Unique Words. Whilst Unique Words as a feature is significant, which was a finding in the original study, when we controlled via normalisation, we found that this was not significant. 
\par
Another observation is that when conducting fitting a linear model to this data does not tell the whole story. For example, consider the following two plots. One shows a model fit to a linear  model, and the other shows a model fit using local polynomial regression model which aims to model the data more closely. It is clear to see that both models can track the movements of language use over time however, the linear model is quite a rigid model which does not allow for different inflection points. It is not the case that language declines in a linear way, as demonstrated by figure  and whilst we have a number of data points that we can use to model RR's decline, in a clinical setting we will often not have this luxury. 
\par

@@

\section{Conclusions}\label{conclusions}
The results of this work show that we can track a person's use of language through time in a number of ways, and it is possible for an individual to be his or her own control. This is important as it means the heterogenous nature of the MCI population does not impact results as much as if we were comparing those with MCI to controls. Equally, it would be helpful to have controls to ascertain what would be usual to expect in the decline of language in a healthy older adult. 
\par 
The results of this work also identify some clues as to what could work in a clinical setting, or how we might be able to collect data in a way that accurately tracks language decline without making incorrect assumptions.
